

%%%%%% PAKKENE TIL CAMBRIDGE %%%%%%%%%%%%%%%%%
  \NeedsTeXFormat{LaTeX2e}[1996/06/01]
\documentclass[prodtf]{EngC}
\usepackage[rightcaption,raggedright]{sidecap}% for side captions
\usepackage{framed}         % for floatingboxes
%			\usepackage{soul}           % for letterspacing in theorem-style headings
\usepackage{soulutf8}
\usepackage[agsm]{harvard} % for the Harvard author-date referencing system
\usepackage[figuresright]{rotating}
\usepackage{floatpag}
\rotfloatpagestyle{empty}
\usepackage[bottom]{footmisc}
\usepackage{amsmath}% if you are using this package, it must be loaded before amsthm.sty
\usepackage{amsthm}
\usepackage{graphicx}

% for multiple indexes using multind.sty
%\usepackage{multind} 
%\ProvidesPackage{multind}
%\makeindex{authors}
%\makeindex{subject}

% for a single index
 \usepackage{makeidx}
 \makeindex
\usepackage{glossaries}
\makeglossaries
\input glossaries.tex


%%% VAARE OPPRINNELIGE PAKKER %%%%%%%%%%%%%%%%%%%%%%%%%
%\usepackage[utf8]{inputenc}
%\usepackage[T1]{fontenc}
%\usepackage{geometry}
%\geometry{a4paper}
%\usepackage[parfill]{parskip}

%%\usepackage{ifpdf}
%%\renewcommand*\familydefault{\sfdefault} %in stead of cmbright
%%\usepackage[cm]{sfmath} %ditto

%\usepackage[slantedGreek]{cmbright}
%\usepackage{amsmath,esint}
%\usepackage{amssymb}
%\usepackage{amstext}
%\usepackage{upgreek}
%\usepackage{ulem}
%\normalem
\usepackage{color}
\usepackage[usenames, dvipsnames]{xcolor}

%\usepackage{booktabs}
%\usepackage{sidecap}
%\usepackage{lineno}
%\usepackage{imakeidx}
%\usepackage{framed}
%\usepackage{caption}
%\makeindex
%\usepackage{soul}
%\usepackage{soulutf8}
%\usepackage{placeins}


%%%%% NOE CAMBRIDGE GREIER JEG IKKE FORSTAAR
  \hyphenation{line-break line-breaks docu-ment triangle cambridge amsthdoc
    cambridgemods baseline-skip author authors cambridgestyle en-vir-on-ment polar}
  \setcounter{tocdepth}{2}% the toc normally lists sections;
% for the purposes of this document, this has been extended to subsections
\makeatletter
\def\TeX{TeX}
\renewcommand{\LaTeX}{La\TeX}
\renewcommand{\LaTeXe}{La\TeX2e}
\makeatother
\raggedbottom

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%% NOEN DEFINISJONER OG VALG
\numberwithin{equation}{chapter}
\numberwithin{figure}{chapter}
\numberwithin{table}{chapter}

\definecolor{newtextcolor}{cmyk}{0.01,0.7,1,0} %% color for new text
\definecolor{newtextcolor2}{cmyk}{1,0.01,0.01,0} %% color for new text
\newcommand{\newtext}[1]{{\color{newtextcolor}#1}}
%\newcommand{\newtextGTE}[1]{{\color{newtextcolor2}#1}}
\newcommand{\BookRef}[1]{{\color{blue}#1}}
\newcommand{\Comment}[1]{{\color{newtextcolor}#1}}
\newcommand{\Figure}[1]{{\color{red}#1}}


%custom color for \hlc
\newcommand{\hlc}[2][yellow]{ {\sethlcolor{#1} \hl{#2}} }
\newcommand{\hlb}[2][blue]{ {\sethlcolor{#1} \hl{#2}} }
\newcommand{\hlr}[2][Maroon]{ {\sethlcolor{#1} \hl{#2}} }
\newcommand{\hlj}[2][OliveGreen]{ {\sethlcolor{#1} \hl{#2}} }
\newcommand{\hlR}[2][red]{ {\sethlcolor{#1} \hl{#2}} }
\newcommand{\hlp}[2][Purple]{ {\sethlcolor{#1} \hl{#2}} }
\newcommand{\hlt}[2][Tealblue]{ {\sethlcolor{#1} \hl{#2}} }
\newcommand{\hlP}[2][Periwinkle]{ {\sethlcolor{#1} \hl{#2}} }


% define \sout{} etc.
\usepackage[normalem]{ulem}

% SI units defines \SI{10}{\kilo\metre}, \si{\newton} etc.
\usepackage{siunitx}
\DeclareSIUnit\molar{M}

% fancyrefs and hyperrefs
\usepackage[plain]{fancyref}
\usepackage[colorlinks = true,
            linkcolor = blue,
            urlcolor  = blue,
            citecolor = blue,
            anchorcolor = blue]{hyperref}

%\newcommand{\ghnote}[1]{{\color{red}#1}}


%boxes and highlight color for text updates, personified!
\newcommand{\ghnote}[1]{\color{white}{\hlb{GH: #1 }}\color{black}}
\newcommand{\ghtxt}[1]{{\color{blue}#1}}
\newcommand{\gtenote}[1]{\color{white}{\hlR{GTE: #1 }}\color{black}}
\newcommand{\gen}[1]{\color{white}{\hlR{GTE: #1 }}\color{black}}
\newcommand{\gtetxt}[1]{{\color{red}#1}}
\newcommand{\gex}[1]{{\color{red}#1}}
\newcommand{\genn}[1]{{\color{orange}#1}}

\newcommand{\orange}[1]{{\color{orange}#1}}
\newcommand{\red}[1]{{\color{red}#1}}
\newcommand{\blue}[1]{{\color{blue}#1}}
\newcommand{\green}[1]{{\color{green}#1}}


\newcommand{\tvnnote}[1]{\color{white}{\hlj{TVN: #1 }}\color{black}}
\newcommand{\tvntxt}[1]{{\color{OliveGreen}#1}}

\newcommand{\snnote}[1]{\color{white}{\hlp{SN: #1 }}\color{black}}
\newcommand{\sntxt}[1]{{\color{Purple}#1}}
\newcommand{\slntxt}[1]{{\color{RoyalPurple}#1}}


\newcommand{\ehnote}[1]{\color{white}{\hlP{EH: #1 }}\color{black}}
\newcommand{\ehtxt}[1]{{\color{Periwinkle}#1}}





\begin{document}
%\linenumbers


%%%%% FRONTMATTER FROM CAMBRIDGE
\frontmatter

%%%%% HALFTITLEPAGE
\begin{halftitlepage}
\halftitle{Book about electrical fields in the brain}
\halfsubtitle{From neurons to extracellular potentials}

%\vskip\baselineskip

\halftitleblurb{This book is about this and that. There will be computational essays. }

%\vskip\baselineskip

\halftitleauthor{Geir Halnes} is me.\\
\halftitleauthor{Torbj{\o}rn V. Ness} is somebody else.\\
\halftitleauthor{Solveig N{\ae}ss} is the third person. \\
\halftitleauthor{Espen Hagen} is number four. \\
\halftitleauthor{Klas H. Pettersen} is not me. \\
\halftitleauthor{Gaute T. Einevoll} is the last author. \\
\end{halftitlepage}

\cleardoublepage

%%%%%%% TITLEPAGE 
\begin{titlepage}
\fmtitle{Book about electrical fields in the brain}
\fmsubtitle{From neurons to extracellular potentials}

\fmtitleauthor{Geir Halnes}
\fmtitleaffil{Norwegian University of Life Sciences, Norway}
\fmtitleauthor{Torbj{\o}rn V. Ness}
\fmtitleaffil{Norwegian University of Life Sciences, Norway}
\fmtitleauthor{Solveig N{\ae}ss}
\fmtitleaffil{University of Oslo, Norway}
\fmtitleauthor{Espen Hagen}
\fmtitleaffil{Norwegian University of Life Sciences, Norway}
\fmtitleauthor{Klas H. Pettersen}
\fmtitleaffil{NORA - The Norwegian Artificial Intelligence Research Consortium, Norway}
\fmtitleauthor{Gaute T. Einevoll}
\fmtitleaffil{Norwegian University of Life Sciences, Norway}

% \cuplogo % This is supposed to insert some cambridge logo which was not included in the template. 
\end{titlepage}


%%%%%%%% COPYRIGHTPAGE 
\begin{copyrightpage}
%\copyrightlogo\par\vskip\baselineskip % % This is supposed to insert some cambridge logo which was not included in the template. 
{\Large Cambridge University Press} \par\vskip1.5\baselineskip
University Printing House, Cambridge CB2 8BS, United Kingdom\par\vskip0.5\baselineskip
One Liberty Plaza, 20th Floor, New York, NY 10006, USA\par\vskip0.5\baselineskip
477 Williamstown Road, Port Melbourne, VIC 3207, Australia\par\vskip0.5\baselineskip
4843/24, 2nd Floor, Ansari Road, Daryaganj, Delhi -- 110002, India\par\vskip0.5\baselineskip
79 Anson Road, \#06-04/06, Singapore 079906\par\vskip\baselineskip

Cambridge University Press is part of the University of Cambridge.\par\vskip0.5\baselineskip

It furthers the University's mission by disseminating knowledge in the pursuit of\par
education, learning, and research at the highest international levels of excellence.\par\vskip\baselineskip

www.cambridge.org\par
Information on this title: www.cambridge.org/9781107164680\par
DOI: 10.1017/9781316691175\par\vskip0.5\baselineskip

\textcopyright\ Geir Halnes et al. 2021 \par\vskip0.5\baselineskip

This publication is in copyright. Subject to statutory exception\par
and to the provisions of relevant collective licensing agreements,\par
no reproduction of any part may take place without the written\par
permission of Cambridge University Press.\par\vskip0.5\baselineskip

First published 2021\par\vskip0.5\baselineskip

Printed in $<$country$>$ by $<$printer$>$\par\vskip0.5\baselineskip

\textit{A catalogue record for this publication is available from the British Library.}\par\vskip0.5\baselineskip

\textit{Library of Congress Cataloging-in-Publication Data}\par\vskip0.5\baselineskip

ISBN 978-1-107-16468-0 Hardback\par
ISBN 978-1-316-61649-9 Paperback\par\vskip0.5\baselineskip

Cambridge University Press has no responsibility for the persistence or accuracy of\par
URLs for external or third-party Internet Web sites referred to in this publication\par
 and does not guarantee that any content on such Web sites is, or will remain,\par
accurate or appropriate.
\end{copyrightpage}


%%%%%%%%% DEDICATIONPAGE
\begin{dedication}
For\\
your eyes only
\end{dedication}


%%%%%%%%%  CONTENT LISTS ETC....
\tocmax{2.10}{2.8.8}{}
\tableofcontents
%\listoffigures
%\listoftables
%\listoffloatingboxes

\cleardoublepage


%%%%%%%% WORK NOTES
\chapter*{Work notes}
\begin{itemize}
\item Structure plan on: https://www.overleaf.com/2551421161hjxkhwwvjrqk
\item Convention: Use $\bf i$ for current density. As Espen noted, it is common to use $\bf J$ for this. However, $\bf J$ is commonly used for fluxes of all sorts. Since we cover electrodiffusion, we need symbols for particle fluxes and flux densities, and I thought it was best to use the $j$'s for that. I also prefer the convention of using the same letter (capitalized or not) for currents and current densities to keep them together:
	\begin{itemize}
	\item $\mathbf{i}$: Current density (A/m$^2$)
	\item $\mathbf{I}$: Current (A)
	\item $\mathbf{j}_k$: Flux density of ion $k$ (mol/m$^2$s)
	\item $\mathbf{J}_k$: Flux of ion $k$ (mol/s)
	\end{itemize}
\gen{Subscripts og superscripts skal ikke vaere i bold.}
\ehnote{jeg vil foreslaa en komplett tabell et sted for symboler brukt gjennom hele denne boken. Ogsaa boer vi vaere obs paa forskjellen mellom skalarfelt og vektorfelt, $J(\mathbf{r})$ vs. $\mathbf{J}(\mathbf{r})$.}
\item Sub or superfix? Suggestion: 
	\begin{itemize}
	\item Superfix to specify types of currents: ${\bf }i^\text{drift}$, ${\bf }i^\text{diff}$, ${\bf }i^\text{cap}$, ${\bf }i^\text{tot}$, ${\bf }i^\text{free}$, ${\bf }i^\text{adv}$.
	\item	Subfix to specify where they are: ${\bf }i_\text{m}$ (membrane), ${\bf }i_\text{t}$ (tissue), ${\bf }i_{n-1,n}$ (between intracellular compartments).
	\item	Subfix to specify ion type: ${\bf }i_\text{Na}$, $i_\text{K}$, $i_{k}$ (general, ion specific), $i_\text{x}$ (general, not ion specific), $i_\text{L}$ (leak).
	\end{itemize}
Disadvantage: HH becomes $i_\text{m} = i^\text{cap} + i_\text{Na} + i_\text{K} + i_\text{L}$ with cap as superfix but the rest as subfix. But maybe that's okay?
\ehnote{Jeg bruker lowercase consistent, i.e., $i_\mathrm{cap}$ -- en kan indeksere som $i_\mathrm{cap}^{\langle i \rangle} \equiv i_\mathrm{cap}[i]$}
\item Potential: $\phi$ or $V$? My favorite is $\phi$. This is greek to Gaute, who seems to prefer $V$. 
\item Refer to main Sections as "Chapters", subsections as Sections? Answer: YES. This is also implicit in the Cambridge-template. 
	\ehnote{Jeg la til en import av fancyref over. Den definerer kommandoen {\textbackslash}Fref\{\} og skal kunne skjoenne forskjellen paa chapter og section utfra innhold i label (chap:*, sec:*, eq:*, etc.), og faa ting internt konsistent. Test:} 
	\ehtxt{{\textbackslash}label\{sec:Intro\}: \Fref{sec:Intro}; {\textbackslash}label\{chap:Intro\}: \Fref{chap:Intro}}
	\ehnote{En del labels er ikke riktig formattert naar det er sagt: {\textbackslash}label\{Basics:eq:UE\} vil ikke fungere for eksempel og boer endres til {\textbackslash}label\{eq:Basics:UE\} e.l.}
\item Reference style: 
	\begin{itemize} 
	\item Harvard: Cambridge gave three choices: Harvard (author?date), Vancouver (numbered),
	and IEEE (numbered). I picked the Harvard-style - I think that is better in a book.
	Unfortunately, Harvard uses the convention that all authors are listed in the first citation and in
	subsequent citations just the first author's name followed by 'et al.' I think that listing all authors
	can get quite messy. This can be overridden by using a duble asterix after \begin{verbatim}	\cite**{} \end{verbatim}.
	\item Basic citation commands are "cite" for (Pettersen et al. 2018) and "citeasnoun" for
	Pettersen et al. (2008).
	\item Urls: I don't know why, but for some reason the harvard-bibtexing did not handle the urls
	in our ".bib"-file, so I removed them all. Now it works fine. I suggest that new bibtex-entries are
	copypasted in from google-scholar. My mendelay-generated entries contained some fields that 	the Harvard-style tried to to things with that it wasnt't able to.
	\ehnote{Hvis du bruker JabRef i nyere versjon og DOI/ISBN er spesifisert, kan du enkelt synke bibtex info -- hoeyreklikk entry saa ligger det litt ned i lista bare}
	\end{itemize}
\item When to include functional arguments and not?
\item Color coding:
	\begin{itemize}
	\item Orange for transmembrane currents: Incscape colorcode: ff7f2aff
	\item Yellow for intracellular currents: Incscape colorcode: ffcc00ff
	\item Red for extracellular currents: Incscape colorcode: ff0c00ff
	\end{itemize}
	\item Nytt fra Espen: Matplotlib-syklus RGBA: (default: cycler('color', '\#1f77b4', '\#ff7f0e', '\#2ca02c', '\#d62728', '\#9467bd', '\#8c564b', '\#e377c2', '\#7f7f7f', '\#bcbd22', '\#17becf')) Legg til alfakode: + 'ff'. Mer info her: \href{https://matplotlib.org/stable/tutorials/colors/colors.html}{https://matplotlib.org/stable/tutorials/colors/colors.html}
\item When to include Units and not?
\item How to make this a "computational essay"?
\item Real booky-books tend to have an Index at the end. Add \verb|\index| around words to add them to Index
\item Boxes as in: https://tex.stackexchange.com/questions/371007/how-to-make-a-box-in-the-sense-of-see-box-1. 
\item SI-pakken % SI units defines \SI{10}{\kilo\metre}, \si{\newton} etc. \usepackage{siunitx}

\end{itemize}
\begin{flushright}\vskip-0.5\baselineskip
%\textit{Ali Woollatt}\\
%\textit{Author locator}
\end{flushright}


%%%%%%%% PREFACE
\chapter*{Preface}
\begin{quote}
\flushright
"An expert is a person who has found out by his own painful experience all the mistakes that one can make in a very narrow field."\\
-- Niels Bohr
\end{quote}
Based on this quote, we judge ourselves to be experts in the narrow field of modelling extracellular potentials.


\def\acrodef#1#2{\noindent\hbox to 38.5pt{\textbf{#1}} #2\par}
\chapter*{Abbreviations}
%\acrodef{WWWOTL}{We Will Work On This List}
%\acrodef{AIWSBC}{And It Will Soon Be Completed}
\acrodef{LFP}{Local Field Potential}
\acrodef{EEG}{...}
\acrodef{ECoG}{...}
\acrodef{MEG}{...}
\acrodef{MUA}{Multiunit Activity}
\acrodef{MEA}{...}
\acrodef{FEM}{...}




\chapter*{Symbols}

\begin{table}[!htb]
\caption{Reserved physical symbols and quantities}
\label{tab:symbols}
\begin{tabular}{|l|l|l|}
\hline
Symbol 				& Unit 					& Description \\
\hline
$t$ 				& \si{\second}  		& time \\
$q$					& \si{\coulomb} 		& electric charge \\
$I, \mathbf{I}$ 				& \si{\ampere} 			& electric current (scalar, directed) \\
$i, \mathbf{i}$ 				& \si{\ampere\per\metre^2} 			& specific electric current (scalar) \\
$E,\mathbf{E}$		& \si{\volt\per\metre} 	& electric field (scalar, directed) \\
$V$ 				& \si{\volt} 			& electric potential \\
$u \in \{ x,y,z \}$	& \si{\metre} 			& location on $u$-axis \\
$\mathbf{r}$		& \si{\metre} 			& position vector \\
\hline
\end{tabular}
\end{table}





\chapter*{Mathematical notation}


\begin{table}[!htb]
\caption{Mathematical notation and operators}
\label{tab:math}
\begin{tabular}{|l|l|}
\hline
Notation/operator & Description \\
\hline
$\coloneq$											& definition operator \\
$\frac{\partial y}{\partial x}$ 					& partial derivative of $y$ with respect to $x$ \\ 
$\nabla \coloneq [ \frac{\partial}{\partial x}, \frac{\partial}{\partial y}, \frac{\partial}{\partial z} ]^T $  & gradient operator \\
$u$ 												& scalar value \\
$u(t)$												& variable/function of $t$ \\
$\langle u(t) \rangle $ 							& mean/expectation value of $u(t)$  \\
$\vert u \vert$										& absolute value of $u$ \\
$\Vert \mathbf{r} \Vert$							& norm of $\mathbf{r}$ \\
$\{ \mathbf{r}_1, \mathbf{r}_2, \ldots \mathbf{r}_n \}$	& set \\
$\mathbf{u} \coloneq [u_1, u_2, \ldots u_n]$ 		& length $n$ vector (1-dimensional matrix) \\
$\mathbf{U} \coloneq [\mathbf{u}_1^T, \mathbf{u}_2^T, \ldots \mathbf{u}_n^T]$ 		& 2-dimensional matrix \\
$\mathcal{U} \coloneq [\mathbf{U}_1, \mathbf{U}_2, \ldots \mathbf{U}_n]$ & 3-dimensional matrix/tensor \\
 & \\
\hline
\end{tabular}
\end{table}

% \editedlistofcontributors
\include{notation}


%%%%%%% END OF FRONTMATTER FROM CAMBRIDGE


\mainmatter

\part{Getting started}
\include{Ch-Intro}
\include{Ch-Basics}

\part{Theory}
\include{Ch-Neuron}
\include{Ch-VC}
\include{Ch-Sigma}
\include{Ch-LFPy}
\include{Ch-Eldiff}

\newpage
\part{Applications}
\include{Ch-Spikes}
\include{Ch-LFP}
\include{Ch-ECoG}
\include{Ch-EEG}
\include{Ch-MEG}
\include{Ch-Stim}

\part{Outlook}
\include{Ch-ephaptic}
\include{Ch-Schemes}
\include{Ch-Tech}
\include{Ch-Last}

\backmatter

\appendix
\chapter{Derivation of four-sphere model}
\tvnnote{Paste from LFPy 2.0 paper?}
\ehnote{Solveig sin artikkel vel? Vi viste ikke denne utledningen i LFPy2.0-paperet.}

\include{AppB-multipole}
%\section{Synapses \Comment{- Gaute}}
%\subsection{Chemical synapses:  postsynaptic response}
%\subsection{Chemical synapses:  synaptic plasticity}
%\subsection{Electrical synapses: gap junctions}
%\section*{References}
%\bibliographystyle{plainnat}
%\bibliographystyle{apalike}

\include{App-linesource}

\endappendix

%%%%% FOR END NOTES: DON'T THINK THAT WE WANT THAT
% insert a blank line to the toc list
%  \addtocontents{toc}{\vspace{\baselineskip}}
%  \theendnotes

\bibliography{ECS_book}\label{refs}
\cleardoublepage

%\section*{Index}
\printindex
\end{document}

